---
title: "NFL Data Exploration"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
### Source: https://www.kaggle.com/maxhorowitz/nflplaybyplay2009to2016
# devtools::install_github(repo = "maksimhorowitz/nflscrapR")
# library(nflscrapR)
library(tidyverse)
library(tidytext)
library(teamcolors)
library(here)
```

```{r Load-Data}
data_set <- read.csv( here( "NFL Play by Play 2009-2018 (v5).csv") )
# glimpse(data_set)
```

```{r Explore-Data-Game-ID-Take-One}
### Goal: To plot winners/losers with some events from the data set.
data_set %>% 
  dplyr::filter( format( as.Date(data_set$game_date), "%Y") == 2009 ) %>% 
  dplyr::filter( game_id == 2009091000 ) %>%
  dplyr::arrange( play_id ) %>%
  dplyr::select( total_home_score, total_away_score, score_differential, game_seconds_remaining, qtr ) %>%
  dplyr::mutate( game_minutes_remaining = round( game_seconds_remaining / 60, 2 ) ) %>%
  dplyr::mutate( game_minutes_elapsed = 60 - game_minutes_remaining ) %>%
  ggplot( aes(x = game_minutes_elapsed) ) +
    geom_line( aes(y = total_home_score, color = "blue"), linetype = "solid" ) +  
    geom_line( aes(y = total_away_score, color = "red"), linetype = "solid" )
### Realized there is a problem with overtime games. 
```



```{r Load-and-Preprocess-Data-Take-Two}
# data_set <- read.csv( here( "data/NFL Play by Play 2009-2018 (v5).csv") ); glimpse(data_set)
nfl_logos_df <- read_csv("https://raw.githubusercontent.com/statsbylopez/BlogPosts/master/nfl_teamlogos.csv")
```

```{r Explore-Data-Game-ID-Take-Two, message = FALSE, warning = FALSE}
library(ggimage)
### Want time elapsed in overtime to add to the 60 minutes of the game 
data_set %>% 
  ### Filter by games in 2009, select the first year, and sort rows by 'play_id'
  dplyr::filter( format( as.Date(data_set$game_date), "%Y") == 2009 ) %>% 
  dplyr::filter( game_id == 2009091000 ) %>%
  dplyr::arrange( play_id ) %>%
  ### Process time remaining variables to be pleasant for plotting
  dplyr::mutate( game_minutes_remaining = round( game_seconds_remaining / 60, 2 ) ) %>%
  dplyr::mutate( game_minutes_elapsed = 60 - game_minutes_remaining ) %>%
  dplyr::mutate( game_minutes_elapsed = case_when( qtr == 5 ~ 15 + game_minutes_elapsed,
                                                   TRUE ~ game_minutes_elapsed ) ) %>%
  ### Add URLs of team logos to data set 
  left_join(nfl_logos_df, by = c("posteam" = "team_code")) %>%
  ### Make ggplot of various data set variables over time. 
  ggplot( aes(x = game_minutes_elapsed) ) +
    # geom_image(aes(image = url, y = home_wp), size = 0.05) +
    geom_line( aes(y = home_wp, color = "blue"), linetype = "solid" ) +  
    geom_line( aes(y = away_wp, color = "red"), linetype = "solid" ) +
    theme_bw() + 
    xlab( "Minutes Played" ) + 
    ylab( "Win Probability" ) + 
    labs( title = "Win Probability Against Time",
          subtitle = "First Game in Data Set" ) 
    # geom_point( y = fourth_down_failed )
```

## Hypothesis Generation 

We can now plot game scores or win probabilities by team against time. We want to add other variables to the plot to examine their effect. Variables of interest include: 

  1. Fourth down conversions
  2. Coach's/replay challenges
  3. Change of possesion 
  4. Incomplete Pass or Interception 
  5. Touchdown (Pass/Rush) - Is a rush/pass touchdown more "valuable" to win probabiity?
  6. Tackled for loss
  7. Sack / Fumble - How often do these lead to changes in possesion? Does it differ by player/team? 
  
Pairs of variables of interest:

  1. Passing plays are characterized by the four-tuple (pass_length, pass_location, air_yards, yards_after_catch). 
  2. Down play conditionals can help characterize how a team runs its offense 
      - Pittsburg: Bell - run on 1^st^ downs; Brown - pass on 3^rd^ downs (plus distribution of pass plays -- short/medium/long)
  3*. Pass player ID and Receiver player ID ( Tom Brady and Randy Moss, Peyton Manning and Marvin Harrison,
                                              Big Ben and Antonio Brown )

There is no play clock variable. 

## ALIN DOING THINGS
```{r}
wins_df = data_set %>% 
  filter( format( as.Date(data_set$game_date), "%Y") == 2009 ) %>% 
  group_by(game_id) %>% 
  summarize(home = first(home_team), 
            away = first(away_team), 
            home_score_final = last(total_home_score),
            away_score_final = last(total_away_score)) %>% 
  mutate(is_tie = ifelse(home_score_final == away_score_final, 1, 0),
         team = ifelse(home_score_final > away_score_final, 
                         home, 
                         away)) %>% 
  count(team, name = 'wins') %>% 
  mutate(losses = 16 - wins,
         win_percentage = round(wins/16,4)) 
# this is missing ties! 
wins_df 
```

```{r}
# need to unstack this shit -.- 
wins_df %>%
  gather(key = result, value = wl, wins:losses) %>% 
  ggplot(aes(x = reorder(team, win_percentage), y = wl, fill = result)) + 
  geom_bar(stat = 'identity') + 
  labs(title ="2009 NFL Team Ranking ", x = "Team", y = "Games Played") +
  coord_flip()
```
```{r}
wins_df %>% 
  ggplot(aes(x = reorder(team, wins), y = wins)) + 
  geom_bar(stat = 'identity') + 
  labs(title ="2009 NFL Team Ranking ", x = "Team", y = "Wins") + 
  coord_flip() + theme_bw()
```

```{r}
# Wrapping our processing into functions! This is important for making reproducible results 
# with changes to certain parameters (good for comparing across time)
yearly_wins <- function(df, year){
  temp_df <- df %>% 
  filter( format( as.Date(data_set$game_date), "%Y") == year ) %>% 
  group_by(game_id) %>% 
  summarize(home = first(home_team), 
            away = first(away_team), 
            home_score_final = last(total_home_score),
            away_score_final = last(total_away_score)) %>% 
  mutate(is_tie = ifelse(home_score_final == away_score_final, 1, 0),
         team = ifelse(home_score_final > away_score_final, 
                         home, 
                         away)) %>% 
  count(team, name = 'wins') %>% 
  mutate(losses = 16 - wins,
         win_percentage = round(wins/16,4)) 
  
  return(temp_df)
}

plt_wins <- function(wins_df){
  plt <- wins_df %>%
  gather(key = result, value = wl, wins:losses) %>% 
  ggplot(aes(x = reorder(team, win_percentage), y = wl, fill = result)) + 
  geom_bar(stat = 'identity') + 
  labs(title ="2009 NFL Team Ranking ", x = "Team", y = "Games Played") +
  coord_flip()
  
  plt
}
```


So now we get a little smarter and remove the year filter so we can facet wrap this shiz

```{r}
wins_df = data_set %>% 
  mutate(year_of_play = format( as.Date(data_set$game_date), "%Y") ) %>% 
  group_by(year_of_play, game_id) %>% 
  summarize(home = first(home_team), 
            away = first(away_team), 
            home_score_final = last(total_home_score),
            away_score_final = last(total_away_score)) %>% 
  mutate(is_tie = ifelse(home_score_final == away_score_final, 1, 0),
         team = ifelse(home_score_final > away_score_final, 
                         home, 
                         away)) %>% 
  count(team, name = 'wins') %>% 
  mutate(losses = 16 - wins,
         win_percentage = round(wins/16,4)) 

wins_df %>%
  gather(key = result, value = wl, wins:losses) %>% 
  ggplot( aes(reorder_within(team, wl, result), win_percentage)) + 
  geom_bar(stat = 'identity') + 
  labs(title ="NFL Team Rankings", x = "Team", y = "Games Played") +
  coord_flip() + 
  facet_wrap(~ year_of_play)
```






```{r Scrap-Code, eval = FALSE}
library(lubridate)
period_to_seconds(ms(test$time))
format( as.POSIXlt("10:00", format="%M:%S"), format = "%M:%S" )
format( as.POSIXlt(test$time, format="%M:%S"), format = "%M:%S" )  
dplyr::transmute( time = period_to_seconds(ms(time)) )

plot( x = test$game_seconds_remaining,
      y = test$total_home_score, 
      type = 'l', col = "blue")
lines( x = test$game_seconds_remaining,
       y = test$total_away_score, 
       type = 'l', col = "red" )
```